---
title: "First-generation Immigrants' Upward Mobility: Exploring the Effects of Language Spoken At Home"
author: "Natania Wong"
date: "2023-10-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("haven")
#install.packages("tidyverse")
#install.packages("readr")
#install.packages('ipumsr')
#install.packages("oaxaca")
#install.packages("quantreg")
#install.packages("kableExtra")
#install.packages("stargazer")
#`install.packages("devtools")
#library(devtools)
#devtools::install_github("strengejacke/sjPlot")
#install.packages("sjPlot")
#install.packages('binsreg')
install.packages("gtsummary")

# Loading libraries
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(tigris)
library(tidycensus)
library(dplyr)
library(readxl)
library(ggpubr)
library(ggthemes)
library(ggplot2)
library(patchwork)
library(MASS)
library(broom)
library(quantreg)
library(Hmisc)
library(kableExtra)
library(broom)
library(stargazer)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(binsreg)
library(gtsummary)
library(texreg)
select <- gtsummary::select
summarize <- dplyr::summarize

## Loading Data
if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("/Users/nataniawong/Desktop/Cal Things/2023-2024/SP24/senior honors thesis/Research/usa_00006.xml")
raw_data <- read_ipums_micro(ddi)
head(raw_data, 5)
```



## Arranging Data
```{r}
## Subset Data: only dating back to 2000
data <- raw_data %>%
  subset(YEAR >= 2000)

## Reformatting data into numeric values
data[,c(28, 29, 31, 32)]=lapply(data[,c(28, 29, 31, 32)], as.numeric)
data[,c(1)]=lapply(data[,c(1)], as.factor)
#data_nonzero[,c(28, 29, 31, 32)]=lapply(data_nonzero[,c(28, 29, 31, 32)], as.numeric)
data$SEX=data$SEX-1

## Exploring Data
# cat("no. of observations: ", dim(data)[1], "\n")
# cat("no. of variables: ", dim(data)[2], "\n")
# cat("names of variables: ", names(data), "\n")
# cat("mean of total income: ", mean(data$INCWAGE), "\n")
# cat("median of total income: ", median(data$INCWAGE), "\n")
```

## Data Manipulation
** Not causal **
In the following code, I will be examining the exiting columns and pass through arguments to mutate new columns.
```{r}
sort(names(data))
# data %>% 
#   select(BPLD) %>% 
#   arrange(BPLD) %>% 
#   filter(BPLD <= 12092) %>% 
#   group_by(BPLD) %>% 
#   count() %>% View()

# Define a function to map the SPEAKENG (english proficiency) values to numeric scale
map_eng_prof <- function(x) {
  case_when(
    x == 1 ~ 1,  # Does not speak English
    x == 6 ~ 2,  # Speaks but not well
    x == 5 ~ 3,  # Speaks well
    x == 4 ~ 4,  # Speaks very well
    x == 3 ~ 5,  # Speaks only English
    TRUE   ~ NA  # Other cases
  )
}

data <- data %>%
  mutate(native = BPLD <= 12092) %>% # Born inside of US: Code 001:120 (BPL)/ Code 00100:12092 (BPLD)
  mutate(citizen_status = CITIZEN==0) %>% ## Citizenship status question is asked for foreign-born, native-born will answer NA
  mutate(speak_eng_home = LANGUAGED == 0100) %>% ## speak_eng_home == 1 if person answers yes
  # mutate(speak_eng_only = SPEAKENG == 3) %>% ## speak_eng_only == 1 if person answers yes
  mutate(eng_proficiency = map_eng_prof(SPEAKENG)) %>% ## mapping function (map_eng_prof)
  mutate(some_college = EDUCD %in% c(065, 070, 071, 080, 090, 100)) %>% ## some college attainment if person has attended some year of college, but has not obtained a degree
  mutate(college_grad = EDUCD %in% c(081:083, 101)) %>% ## person has obtained a degree
  mutate(post_grad = EDUCD %in% c(081:083, 101, 114:116)) ## person has obtained a post-grad degree

map_race <- function(x) {
  case_when(
    x == 1 ~ "White",  
    x == 2 ~ "Black/African American",  
    x == 3 ~ "American Indian or Alaska Native",  
    x == 4 | 5 | 6 ~ "Asian or Pacific Islander",  
    x == 7 | 8 | 9 ~ "Other races or mixed race"
  )
}

# Add new column (word version of race)
data$agg_race = map_race(data$RACE)
data_cleaned <- data %>% 
  subset(INCWAGE > 0 & INCWAGE < 999999)

#data$native <- factor(data$native,
#                      levels = c("TRUE", "FALSE"))
#data$speak_eng_home <- factor(data$speak_eng_home,
#                              levels = c("TRUE", "FALSE"))
```

## Data exploration
```{r}
## Distribution of citizenship status
# p3 <- data%>%
#   ggplot(aes(x=factor(CITIZEN))) +
#   geom_bar() +
#   theme_bw() 
# 
# p4 <- data%>%
#   ggplot(aes(x=factor(citizen_status))) +
#   geom_bar() +
#   theme_bw() 
# 
# p3 | p4
```


```{r}
# data%>%
#   ggplot(aes(x=log(INCWAGE))) +
#   geom_histogram() +
#   theme_bw() 
```



```{r}
## Proportion of people who do not have an income
# no_wage <- data[data$INCWAGE==0,]
# prop_no_wage <- nrow(no_wage) / nrow(data)
# cat("prop of no wage: ", prop_no_wage, "\n")
# 
# no_totincome <- data[data$INCTOT==0,]
# prop_no_totwage <- nrow(no_totincome) / nrow(data)
# cat("prop of no total wage: ", prop_no_totwage, "\n")
```

For the purpose of this study, I will subset the population that is not in the labor force, since they would not reflect the results of the topic in interest. I will also subset the results to only those with a positive wage, as well as removing the outliers who have a salary of 999999, since that is a representation of NA.
```{r}
## Subset Data: only those in labor force, and those who do not have a wage of 999999
# data <- data %>%
#   subset(LABFORCE!=1, EMPSTAT!=3)%>% 
#   subset(INCWAGE < 999999)
# 
# # INCWAGE > 0 (control shift c)
# 
# cat("prop of no wage (after subsetting): ", 
#     nrow(data[data$INCWAGE==0,]) / nrow(data), "\n")
# 
# cat("prop of no total wage (after subsetting): ", 
#     nrow(data[data$INCTOT==0,]) / nrow(data), "\n")
# 
# ## Non-citizen wage distribution (after subsetting)
# non_native <- data[data$native==0,]
# nc1 <- non_citizen%>%
#   ggplot(aes(x=INCWAGE)) +
#   geom_histogram() +
#   theme_bw() +
#   labs(title = "Non-citizen Wage Distribution (After subsetting)")
# 
# ## Citizen wage distribution (after subsetting)
# citizen <- data[data$citizen_status==1,]
# c1 <- citizen%>%
#   ggplot(aes(x=INCWAGE)) +
#   geom_histogram() +
#   theme_bw() +
#   labs(title = "Citizen Wage Distribution (After subsetting)")
# 
# nc1 | c1

nrow(data[data$INCWAGE==0,])/nrow(data)
data_nonzero <- data %>% 
  subset(INCWAGE != 0)
hist(log(data_nonzero$INCWAGE))
data_nonzero%>%
  ggplot(aes(x=INCWAGE)) +
  geom_histogram() +
  theme_bw()
```


```{r}
s1 <- data %>%
  group_by(YEAR, native, speak_eng_home) %>%
  summarize(mean_salary = mean(INCWAGE)) %>%
  ggplot(aes(x = YEAR, y = mean_salary, color = speak_eng_home)) +
  geom_line() +
  labs(title = "Mean Salary Over the Years By Nativity",
       x = "Year",
       y = "Mean Salary") +
  theme_bw() +
  facet_wrap(~native)


s2 <- data %>%
  group_by(YEAR, native, speak_eng_home) %>%
  summarize(mean_salary = mean(INCWAGE)) %>%
  ggplot(aes(x = YEAR, y = mean_salary, color = native)) +
  geom_line() +
  labs(title = "Mean Salary Over the Years By Speaking Eng at Home",
       x = "Year",
       y = "Mean Salary") +
  theme_bw() +
  facet_wrap(~speak_eng_home)

s1/s2
```



```{r}
# Distribution of people who speak english at home
# data%>%
#   ggplot(aes(x=speak_eng_home)) +
#   geom_bar() +
#   theme_bw() 
# 
# # Distribution of people's self-assessed english ability
# data%>%
#   ggplot(aes(x=factor(SPEAKENG))) +
#   geom_bar() +
#   theme_bw() 
```


## OLS Regression

```{r}

data$log_INCWAGE <- log(data$INCWAGE)

#lm
lm0 = lm(INCWAGE  ~ native + speak_eng_home, data=na.omit(data))
summary(lm0)

lm0_tidy <- tidy(lm0) 
lm0_glance <- glance(lm0) 

# Extract relevant information
r_squared <- lm0_glance$r.squared
aic <- lm0_glance$aic
bic <- lm0_glance$bic
log_lik <- lm0_glance$logLik

# Create a data frame for additional information
additional_info <- data.frame(
  Metric = c("R-squared", "AIC", "BIC", "Log.Lik."),
  Value = c(r_squared, aic, bic, log_lik)
)

# Combine with the original table
combined_table <- bind_rows(lm0_tidy, additional_info)

# Print the combined table
kbl(lm0_tidy, caption = "Table 2: OLS Regression Model 1 Results", format = "html", align = "r") %>%
  kable_classic(full_width = F, html_font = "helvetica") 

vars <- c("INCWAGE")
sumtable(data, var=vars)

summary(data$INCWAGE)

```


```{r}
lm1 = lm(INCWAGE ~ native + AGE + SEX + speak_eng_home + some_college + college_grad + post_grad, data=data)
summary(lm1)

tidy(lm1) %>%  
  kbl(caption="Table 3: OLS Regression Model 2 Results",
      format= "html",
      align="r") %>%
  kable_classic(full_width = F, html_font = "helvetica") 

```

```{r}
lm2 = lm(INCWAGE ~ native * speak_eng_home + AGE + SEX + some_college + college_grad + post_grad, data=data)
summary(lm2)

#quantile regression
rq0 <- rq(INCWAGE ~ native + AGE + SEX + speak_eng_home + some_college + college_grad + post_grad, data = data, weights = HHWT, method = "fn", tau = 0.5)
summary(rq0)

rq1 <- rq(INCWAGE ~ native * speak_eng_home + AGE + SEX + some_college + college_grad + post_grad, data = data, weights = HHWT, method = "fn", tau = 0.5)
summary(rq1)
```


## ALternative OLS (By grouping salary)
```{r}
quantile(data[data$YEAR==2000,]$INCWAGE)
quantile(data[data$YEAR==2006,]$INCWAGE)
quantile(data[data$YEAR==2011,]$INCWAGE)
quantile(data[data$YEAR==2016,]$INCWAGE)

# # 2000, Wages: 75th percentile 
# lm(INCWAGE ~ native + AGE + SEX + citizen_status + speak_eng_home + some_college + college_grad + post_grad, data=data[data$YEAR==2000 & data$INCWAGE >= 40000,])
# 
# # 2006, Wages: 75th percentile 
# lm(INCWAGE ~ native + AGE + SEX + citizen_status + speak_eng_home + some_college + college_grad + post_grad, data=data[data$YEAR==2006 & data$INCWAGE >= 50000,])
# 
# # 2011, Wages: 75th percentile 
# lm(INCWAGE ~ native + AGE + SEX + citizen_status + speak_eng_home + some_college + college_grad + post_grad, data=data[data$YEAR==2011 & data$INCWAGE >= 55000,])
# 
# # 2016, Wages: 75th percentile 
# lm(INCWAGE ~ native + AGE + SEX + citizen_status + speak_eng_home + some_college + college_grad + post_grad, data=data[data$YEAR==2016 & data$INCWAGE >= 63000,])
```

## Blinder-Oaxaca Decomposition
```{r}
library(oaxaca)

results0 <- oaxaca(INCWAGE ~ AGE + SEX  + speak_eng_home + some_college + college_grad + post_grad | native, data=data)

results0$n # Group B: data$native == 1
results0$y

results0$twofold$overall

plot(results0, decomposition = "twofold", group.weight = -1)
 
```

```{r}
tidy(results0) %>%  
  kbl(caption="Table 3: OLS Regression Model 2 Results",
      format= "html",
      align="r") %>%
  kable_classic(full_width = F, html_font = "helvetica") 
```




```{r}
no_postgrad <- data %>%
  subset(post_grad!=1)

results1 <- oaxaca(INCWAGE ~ AGE + SEX  + speak_eng_home + some_college + college_grad | native, data=no_postgrad)

results1$n # Group B: data$native == 1
results1$y

results1$twofold$overall

plot(results1, decomposition = "twofold", group.weight = -1)
```

